---
title: "Reproduce Figure 1a as it Appears in the Manuscript"
author: "Samuel Sarria Hurtado and Paul Sheridan"
output: html_notebook
---

```{r, include = FALSE}
# Delete all variables initialized in R session, if any
rm(list = ls())

# Load libraries
library(dplyr)
library(here) # for working with file paths
library(rjson) # for reading JSON files
library(magicaxis) # for prettying up plot
library(scales) # for custom colors

# Set directory relative paths
data_dir <- file.path(here(), "genia/1-preprocessing")
output_dir <-file.path(here(), "genia/3-figures/figure-1")

# Set random seed
set.seed(86472)

# Define custom colors
hex <- hue_pal()(3)
gg_red <- hex[1] # ggplot default red
gg_green <- hex[2] # ggplot default green
gg_blue <- hex[3] # ggplot default blue
```


# Overview

The code in this R Markdown file reproduces the plot of Figure 1a from the manuscript "A heuristic approach to term dispersion quantification with applications to domain-specific terminology extraction" by Sarria Hurtado et al. (2026).


# Read and Preprocess the Data

Read GENIA Term corpus and preprocess the documents therein for downstream analysis.

```{r}
# Read data
infile <- file.path(data_dir, "GENIAcorpus3.02-preprocessed.json")
raw_doc_strings <- fromJSON(paste(readLines(here(infile)), collapse = ""))

# Convert raw doc strings to tokens
d <- length(raw_doc_strings)
raw_docs <- vector("list", length = d)

for (j in 1 : d) {
  raw_docs[[j]] <- strsplit(raw_doc_strings[[j]], split = " +")[[1]]
}
```


# Word Statistics

Collect basic word statistics used to generate the plot.

```{r}
# Calculate basic stats
d <- length(raw_docs)
nij <- vector("list", length = d)
doc <- vector("list", length = d)
all_terms_table <- table(unlist(raw_docs))
all_terms <- names(all_terms_table)
all_term_counts <- as.numeric(all_terms_table)

for (j in 1 : d) {
  mydoc <- raw_docs[[j]]
  mytable <- sort(table(mydoc), decreasing = T)
  nij[[j]] <- as.numeric(mytable)
  doc[[j]] <- names(mytable)
}

nj <- unlist(lapply(nij, sum))
bj <- unlist(lapply(nij, length))
term <- unique(unlist(doc))
m <- length(term)
n <- sum(unlist(nij))
ni <- table(unlist(raw_docs))[term]
bi <- table(unlist(doc))[term]
DP <- as.numeric(bi / d)
CP <- as.numeric(ni / n)
IDF <- as.numeric(log(d / bi))
ICF <- as.numeric(log(n / ni))
```



# Calculate Expected Relations

Calculate the expected relations of Eqs. (1) and (2).

```{r}
# Calculate expected relations
A <- min(CP)
B <- 0.07
theta_theoretical <- exp(seq(from = log(A), to = log(B), by = 1 / 100))
no_of_theta_theoretical_values <- length(theta_theoretical)
EDP <- numeric(no_of_theta_theoretical_values)
ECP <- numeric(no_of_theta_theoretical_values)
for (i in 1 : no_of_theta_theoretical_values) {
  ECP[i] <- theta_theoretical[i]
  EDP[i] <- 1
  for (j in 1 : d) {
    EDP[i] <- EDP[i] - (1 / d) * (1 - theta_theoretical[i]) ^ nj[j]
  }
}

EICF <- (1 - theta_theoretical) / (2 * n * theta_theoretical) - log(theta_theoretical)

EIDF <- numeric(no_of_theta_theoretical_values)
for (i in 1 : no_of_theta_theoretical_values) {
  C0 <- 0
  for (j in 1 : d) {
    C0 <- C0 + (1 - theta_theoretical[i]) ^ nj[j] - (1 - theta_theoretical[i]) ^ (2 * nj[j])
  }
  EIDF[i] <- C0 / (2 * d ^ 2 * EDP[i] ^ 2) - log(EDP[i])
}

IDF_runner <- log(d / seq(max(bi), 1))
ICF_upper_bound <- as.numeric(log(n / seq(max(bi), 1)))
```




# Generate Scatterplot

Estimate slope and intercept for the asymptotic relation of Eq. (2).

```{r}
# Calculate parameters for plotting asymptotic relation
lm_fit <- lm(ICF_upper_bound ~ IDF_runner)
intercept <- coef(lm_fit)[1]
slope <- coef(lm_fit)[2]
```


Deduplicate points for plotting.

```{r}
# Deduplicate
mydata <- data.frame(IDF = IDF, ICF = ICF) |> distinct()
IDF <- mydata$IDF
ICF <- mydata$ICF
```


Generate the scatterplot and save to PDF.

```{r}
# Plot
outfile <- file.path(output_dir, "figure-1a.pdf")
pdf(here(outfile), width = 6, height = 6, useDingbats = FALSE)
  plot(x = IDF,
    y = ICF,
    main = "GENIA Term corpus",
    xlab = "IDF",
    ylab = "ICF",
    axes = FALSE,
    type = "n")
  grid(lwd = 2)
  magaxis()
  points(x = IDF, y = ICF, pch = 16, cex = 0.75)
  points(x = EIDF, y = EICF, type = "l", lwd = 3, col = gg_red)
  abline(a = intercept, b = slope, lty = "dashed", lwd = 3, col = gg_blue)
  legend("topleft",
       legend = c("Expected relation of Eq. (1)", "Asymptotic relation of Eq. (2)"),
       inset = 0.01,
       col = c(gg_red, gg_blue),
       lty = c(1, 2),
       cex = 0.8,
       lwd = 3,
       seg.len = 3,
       bty = "n")
dev.off()
```




